\chapter{Addressing Energy Distribution with IDEA}
\label{chapter-idea}

Building on the core ideas developed by Lance, IDEA (Integrated Distributed
Energy Awareness) takes two new directions. First, we try and build a true
system service that multiple components can leverage, rather than focusing on
a single energy-intensive component (bulk data transfer) as was done in
Lance. Existing sensor network platforms provide little support for
collaborative energy management. Approaches such
TinyOS~\cite{tinyos-asplos00}, Pixie~\cite{pixie-sensys08},
Eon~\cite{eon-sensys07} and Levels~\cite{levels-sensys07} facilitate local
control only, failing when greedy node energy minimization fails to produce
the best outcome. Network-wide solutions such as
Lance~\cite{lance-sensys08}, Mercury~\cite{parkinsons-embs07}, and
EnviroMic~\cite{enviromic} either require centralized control or are tailored
to the needs of a specific application domain. In sensor networks the
majority of energy consumption is consumed by multi-node collaboration. We
argue that due to the distributed nature of energy consumption and
availability, improving performance requires consideration of both
\textit{where} energy is and \textit{how much} is being used.

Second, we attempt to address the core challenge of unequal energy
distribution within the sensor network itself, particularly as it effects the
overall data quality delivered by the macroscope. Energy-harvesting sensor
networks experience variations in load and charging rates that threaten
high-fidelity operation. Changing application demands can produce variations
in load rates, while energy-harvesting properties can produce variations in
charging rates. Energy mismanagement can lead to reduced fidelity, when
nodes' batteries empty, or wasted energy, when nodes harvest energy they
cannot store.

Energy harvesting capabilities such as solar charging further complicate the
distributed energy management task. The energy collected at each node may
vary significantly based on node placement, and the energy collected
day-to-day may vary significantly based on weather patterns. Preparing the
network for overnight operation requires capturing as much energy as possible
during the day and minimizing energy wasted by charging full batteries, while
overnight operation itself requires adjusting the network's load profile to
match the distribution of energy stored during daytime.

Fortunately, dense networks provide redundancy that can be used to control
the distribution of energy usage. Multiple possible routing paths may
connect a node to the sink. Tuning MAC parameters allows nodes to shift
communication load to their neighbors. Sensor inputs from multiple nodes may
be redundant, allowing some to be disabled or operated at reduced fidelity.
The existence of these choices implies that it is possible to tune the energy
load of the network to better match energy availability. Effective load
tuning can increase the fidelity provided to the application at a fixed
battery size, or allow battery sizes to be reduced while maintaining the
required fidelity level.

Matching load to availability across the network requires
\textit{integrating} with application components producing energy load,
\textit{distributing} load and availability information to facilitate node
decision making, and \textit{awareness} of the connection between load,
availability, and application-level fidelity. We propose Integrated
Distributed Energy Awareness (IDEA), a sensor network service addressing
these goals. IDEA monitors and models the load and charge rates on each node.
To allow nodes to reason about their impact on others, each node distributes
its model parameters, updating them as necessary to ensure continued
accuracy. IDEA clients are responsible for estimating their own distributed
energy impact. When changing state, IDEA helps them evaluate each proposed
option using an energy objective function tailored to meet specific
application goals. By tracking availability and informing the energy
decision-making process, IDEA simplifies the construction of energy-aware
components.

The rest of this chapter is structured as follows:

\begin{enumerate}

\item Section~\ref{idea-sec-motivation} motivates the IDEA approach a simple
example.

\item In Section~\ref{idea-sec-architecture}, we describe IDEA, a new service
uniting energy monitoring, load modeling and distributed state sharing into a
single service facilitating distributed decision making.

\item We present three case studies illustrating how to use IDEA in
Section~\ref{idea-sec-casestudies}, including an existing routing protocol
modified to choose energy-aware routes, and a application using IDEA to
determine how to localize acoustic events.

\item Using simulation and testbed results, in
Section~\ref{idea-sec-evaluation} we compare the performance of IDEA with
approaches that do not consider energy distribution, showing that IDEA
enables improvements in lifetime of up to 27\%.

\item Finally, we review related work in Section~\ref{idea-sec-related}.

\end{enumerate}
