\section{Introduction}

% Sensor networks used in many scientific explorations:
%     - Habitat monitoring (GDI, original Cerpa paper)
%     - Environment (redwoods) soil monitoring
%     - Civil structures (NetSHM, Wisden, GGB)
% Despite the great interest, however, not much has been done to
% validate the sensor network's ability to provide scientifically
% meaningful data to end-users. A number of challenges arise in
% sensor nets that make this difficult: sensor calibration, node
% failure, message loss, and imprecise time synchronization.
% However, if sensor nets are ultimately going to be successful
% in scientific field work they must be held to a very high standard;
% meeting or exceeding the capabilities of existing scientific 
% instrumentation.

Sensor networks are making inroads into a number of scientific explorations,
including environmental monitoring~\cite{rope-emnets05,berkeley-redwoods},
habitat monitoring~\cite{cerpa-habitat,mainwaring-habitat,gdi-sensys04}, and
structural monitoring~\cite{ggb-monitoring,netshm-emnets05,wisan}.  In each
of these domains, the use of low-power wireless sensors offers the potential
to collect data at spatial and temporal scales that are not feasible with
existing instrumentation.  Despite increased interest in this area, little
has been done to evaluate the ability of sensor networks to provide
meaningful data to domain scientists. A number of challenges confound such an
effort, including node failure, message loss, sensor calibration, and
inaccurate time synchronization. To successfully aid scientific studies,
sensor networks must be held to the high standards of 
%meeting or exceeding 
current scientific instrumentation.
%capabilities.

% In this paper, we take a hard
% look at the performance of a sensor network deployed on an active
% volcano with a focus on its effectiveness as a scientific instrument.
% This requires evaluation along two core metrics: data fidelity
% and data yield. Fidelity is defined in terms of the quality of
% the captured signals both in terms of individual samples as well
% as consistency of the signals across multiple nodes. Data yield is
% a measure of the quantity of data returned by the network during
% the field deployment.

In this paper, we take a hard look at the performance of a wireless sensor
network deployed on an active volcano. We evaluate its effectiveness as a
scientific instrument using two metrics: {\em data fidelity} and {\em yield}.
Data fidelity encompasses the quality and consistency of retrieved
seismoacoustic signals, while data yield reflects the quantity of data
delivered by the network. 
%and latencies associated with data collection.  the network delivered.

% Monitoring volcanic eruptions raises new challenges for sensor
% networking as the data fidelity requirements are very high.
% First, data rates are relatively high with each node sampling
% multiple channels of data at 100 Hz. This is in contrast to
% many previous sensor network applications that focus on
% very low-rate data collection (one sample every few minutes).
% Second, signal collection must be reliable with no dropped samples
% in order to support detailed analysis of the signal. Third,
% data from multiple nodes must be accurately timestamped against
% a global clock (typically tied to GPS) to support comparisons of
% signals across multiple nodes; e.g., determining the seismic
% wave arrival.

Typical volcano monitoring studies employ GPS-synchronized data loggers
recording both seismic and acoustic signals. These provide high data fidelity
and yield but are bulky, power hungry, and difficult to deploy. Existing
analog and digital telemetry is similarly cumbersome. 
% MDW: The spin here is CRITICAL: we are not just replacing dataloggers with
% motes.
The use of wireless sensors could enable studies involving many more sensors
distributed over a larger area. However, the science requirements pose a
number of difficult challenges for sensor networks.  First, seismoacoustic
monitoring requires high data rates, with each node sampling multiple
channels at 100~Hz. Second, signal analysis requires complete data,
necessitating reliable data collection. Third, volcano studies compare
signals across multiple sensors, requiring that collected data be accurately
timestamped against a GPS-based global clock.
% MDW: "meaningful global clock" is too vague.

% The core contribution of this paper is a thorough analysis of the
% efficacy of a volcano-monitoring sensor network as a scientific
% instrument, and is the first paper (to our knowledge) to take a
% holistic view of sensor network performance and data quality against
% such a stringent measure. XXX Check this claim against GDI stuff!!!

% MDW: "core" makes it stronger. 
The core contribution of this paper is an analysis of the efficacy and
accuracy of a volcano-monitoring sensor network as a scientific instrument.
This is the first paper to our knowledge to take a science-centric view of a
sensor network with such demanding data-quality requirements.  In this paper,
we evaluate the data collected from a 19-day field deployment of 16~wireless
sensors on Reventador volcano, Ecuador, along the following axes:

\begin{list}{$\bullet$}{\setlength{\topsep}{0.05in}
                        \setlength{\leftmargin}{0.15in}
                        \setlength{\itemsep}{0.05in}}
%\begin{itemize}{}{\setlength{\topsep}{1.0in}}
%   - Robustness: How effectively did the network continuously 
%     report data during its lifetime. We find that the sensor
%     nodes themselves were very reliable (with uptimes exceeding
%     99%) but the base station infrastructure suffered numerous
%     outages due to power failures.

\item 
{\bf Robustness:} We find that the sensor nodes themselves were extremely
reliable but that overall robustness was limited by power outages at the base
station and a single three-day software failure. Discounting the power
outages and this single failure, mean node uptime exceeded 96\%.
%
%   - Event detection accuracy: Our network was designed to
%     trigger data collection following earthquakes and eruptions
%     and we evaluate the effectiveness of our distributed event
%     detector in capturing signals of interest.
%
% MDW: "interesting seismoacoustic events" is jargon and too vague.
\item 
{\bf Event detection accuracy:} Our network was designed to trigger data
collection following volcanic events such as earthquakes and eruptions. We
measure the accuracy of our distributed event-detection algorithm, finding
that the algorithm has a zero false positive rate. However, the network
failed to detect many seismic events due to a poor choice of event-detection
parameters and limitations of our data collection protocol.
%Establishing detection 
%accuracy is hindered by the lack of complete ground truth.
%
%
%   - Performance: We evaluate the ability of the network to deliver
%     complete signals to the base station following an event, 
%     both in terms of the yield as well as throughput and latency.
%
\item 
{\bf Data transfer performance:} We evaluate the ability of our data
collection protocol to transfer complete signals following an event. We find
a 90th percentile {\em event yield} (fraction of nodes for which all data for
an event was collected) of 94\% and a latency of 63~sec per radio hop for
downloading 60~sec worth of data.

%   - Timing accuracy: As mentioned above, time synchronization is
%     essential. We evaluate both the raw performance of the
%     underlying time synchronization protocol (FTSP) and describe
%     a new technique for time rectification of the captured signals
%     allowing us to map each sample onto a global, GPS-based
%     timestamp despite only having a single GPS receiver in the field
%     and failures of the FTSP protocol.
%
\item 
{\bf Timing accuracy:} Data collected by each node must be timestamped to
within a single sample time (10~ms) to enable seismological analysis. We
evaluate the stability of the underlying time synchronization protocol
(FTSP~\cite{ftsp}), and develop a novel approach to {\em time rectification}
that accurately timestamps each sample despite failures of the FTSP protocol.
We show that this approach recovers timing with a 90th-percentile error of
6.8~msec in a 6-hop network.
%
%   - Data fidelity: Finally, we take a seismological view of the
%     captured data and present a head-to-head comparison of signals
%     recorded by our sensor network against a data logger colocated
%     with one of our nodes; and also look at the consistency of the
%     data across the network, with an interesting seismological
%     finding that suggests many earthquakes originated near the
%     {\em middle} of our sensor array rather than at or near the
%     vent.
%
\item 
{\bf Data fidelity:} Finally, we take a seismological view of the captured
data and present a head-to-head comparison of data recorded by our sensor
network against a colocated data logger.  We also evaluate the consistency of
the recorded signals in terms of seismic and acoustic wave arrival times
across the network, showing that the data is consistent with expected
physical models of the volcano's activity.
% MDW: I think the discrimination is irrelevent here.
%which are used to
%discriminate different types of volcanic activity.
%
%\end{itemize}
\end{list}

The rest of this paper is organized as follows. The next section provides
background on the use of wireless sensors for volcano monitoring and outlines
the underlying science goals. In Section~\ref{evaluation-sec-design} we
describe the architecture of our system and the field deployment at
Reventador.
Sections~\ref{evaluation-sec-robustness}~through~\ref{evaluation-sec-fidelity}
present a detailed analysis of the network's performance along each of the
evaluation metrics described above. Section~\ref{evaluation-sec-related}
discusses related work and Section~\ref{evaluation-sec-lessons} presents
several lessons learned from the deployment.
Section~\ref{evaluation-sec-conclusions} outlines future work and concludes.

