\documentclass[lettersize]{sig-alternate-konrad}
\usepackage{listings}
\usepackage{times}  % Use Adobe Times font set
%\usepackage{epsfig,twocolumn}
\usepackage{url}
\usepackage{textcomp}
\usepackage{graphicx} 
\usepackage{color}
\usepackage{multirow}

% This will generate links within the PDF as well as a set of bookmarks
% \usepackage[colorlinks,pdfpagemode=None]{hyperref} 

%% ----- Customizations -----
\setlength{\abovecaptionskip}{-0.05in}
\setlength{\belowcaptionskip}{-0.05in}

\newcommand{\XXXnote}[1]{{\bf\color{red} XXX: #1}}
%\newcommand{\XXXnote}[1]{}

\pagestyle{empty}
%\pagenumbering{arabic}

\begin{document}

%%%%%%%%%%%% THIS IS WHERE WE PUT IN THE TITLE AND AUTHORS %%%%%%%%%%%%

\conferenceinfo{SenSys'08,} {November 5--7, 2008, Raleigh, North Carolina, USA.} 
\CopyrightYear{2008}
\crdata{978-1-59593-990-6/08/11}

\title{{\ttlfnt Lance: Optimizing High-Resolution Signal Collection
in Wireless Sensor Networks}}
\numberofauthors{1}
\author{\alignauthor Geoffrey Werner-Allen$^\dag$, Stephen
Dawnson-Haggerty$^\star$ and Matt Welsh$^\dag$\\
{\affaddr $^\dag$ School of Engineering and
Applied Sciences, Harvard
University, Cambridge, MA 02138} \\
{\affaddr $^\star$ University of California, Berkeley, Berkeley, CA 94720} \\
\email{werner@eecs.harvard.edu, stevedh@cs.berkeley.edu,
mdw@eecs.harvard.edu}
{\affaddr \texttt{http://fiji.eecs.harvard.edu/Lance}}}
\maketitle

\maketitle
\thispagestyle{empty}

%%%%%%%%%%%%%  ABSTRACT GOES HERE %%%%%%%%%%%%%%

\subsection*{Abstract}
An emerging class of sensor networks focuses on reliable 
collection of high-resolution signals from across the network.
In these applications, the system is capable of acquiring more 
data than can be delivered to the base station, 
due to severe limits on radio bandwidth and energy.
Moreover, these systems are unable 
to take advantage of conventional approaches to in-network data
aggregation, given the high data rates and need for raw signals.
These systems face an important challenge: how to
maximize the overall value of the collected data, subject to resource constraints.

In this paper, we describe {\em Lance}, a general approach to
bandwidth and energy management for reliable data collection in wireless
sensor networks. Lance couples the use of optimized, data-driven 
reliable data collection with a model of energy cost for extracting 
data from the network. Lance's design decouples resource allocation 
mechanisms from application-specific policies, enabling flexible 
customization of the system's optimization metrics.

We describe the Lance architecture in detail, demonstrating its
use through a range of target applications and resource management
policies. We present an extensive study driven by both 
real and synthetic data distributions, through simulations and runs
on a large sensor testbed.  We show that Lance maximizes 
the value of the collected data under a range of resource 
constraints, achieving near-optimal allocation of radio bandwidth and
energy. Finally, we present results from a real sensor network
deployment at Tungurahua volcano, Ecuador, in which Lance was used to
drive data collection for an eight-node network collecting seismic and
acoustic signals from the active volcano.  

%%%%%%%%%%%%%  BODY OF PAPER GOES HERE %%%%%%%%%%%%%%


\category{C.2.4}{Distributed Systems}{}
\category{C.3}{Special-Purpose and Application-Based
Systems}{Real-time and embedded systems}
\vspace{-0.1in}
\terms{Design}
\vspace{-0.1in}
\keywords{Wireless Sensor Networks, Data Collection, Resource
Management}

\input{intro.tex}
\input{background.tex}
\input{architecture.tex}
\input{policies.tex}
\input{applications.tex}
\input{implementation.tex}
\input{evaluation.tex}
\input{deployment.tex}
\input{conclusions.tex}

%%%%%%%%%%%%%  THIS IS WHERE THE BIBLIOGRAPHY GOES %%%%%%%%%

%\def\baselinestretch{0.92}
%\begin{footnotesize}
%\setlength{\itemsep}{1in}
\bibliographystyle{abbrv} \bibliography{sensys08}
%\end{footnotesize}

\end{document}
