\chapter{Evaluation of 2005 Deployment}
\label{chapter-evaluation}

In this chapter, we take a hard look at the performance of a wireless sensor
network deployed on an active volcano. We evaluate its effectiveness as a
scientific instrument using two metrics: data \textit{fidelity} and
\textit{yield}. Fidelity encompasses the quality and consistency of retrieved
signals, while yield reflects the quantity of data delivered by the network.

We begin by describing the architecture
(Section~\ref{evaluation-sec-architecture}) and deployment
(Section~\ref{evaluation-sec-deployment}) of a volcano-monitoring sensor
network. Our system was deployed on Reventador volcano in Ecuador and ran for
19~days. The core contribution of this chapter is an analysis of the efficacy
and accuracy of a volcano-monitoring sensor network as a scientific
instrument. The rest of the chapter evaluates our field deployment along the
following axes:

\begin{itemize}

\item \textbf{Network robustness:} In Section~\ref{evaluation-sec-robustness}
we find that the sensor nodes themselves were extremely reliable but that
overall robustness was limited by power outages at the base station and a
single three-day software failure. Discounting the power outages and this
single failure, mean node uptime exceeded 96\%.

\item \textbf{Event detection accuracy:} Our network was designed to trigger
data collection following volcanic events such as earthquakes and eruptions.
In Section~\ref{evaluation-sec-eventdetection} we measure the accuracy of our
distributed event-detection algorithm, finding that the algorithm has a zero
false positive rate. However, the network failed to detect many seismic
events due to a poor choice of event-detection parameters and limitations of
our data collection protocol.

\item \textbf{Data collection performance:}
Section~\ref{evaluation-sec-performance} evaluates the ability of our data
collection protocol to transfer complete signals following an event. We find
a 90th percentile \textit{event yield} (fraction of nodes for which all data
for an event was collected) of 94\% and a latency of 63~s per radio hop for
downloading 60~s worth of data.

\item \textbf{Timing rectification and accuracy:} Data collected by each node
must be timestamped to within a single sample time (10~ms) to enable
seismological analysis. Section~\ref{evaluation-sec-timing} evaluates the
stability of the underlying time synchronization protocol (FTSP~\cite{ftsp}),
and presents a novel approach to \textit{time rectification} that accurately
timestamps each sample despite failures of the FTSP protocol. We show that
this approach recovers timing with a 90th-percentile error of 6.8~ms in a
6-hop network.

\item \textbf{Data fidelity:} In Section~\ref{evaluation-sec-fidelity} we
take a seismological view of the captured data and present a head-to-head
comparison of data recorded by our sensor network against a colocated data
logger. We also evaluate the consistency of the recorded signals in terms of
seismic and acoustic wave arrival times across the network, showing that the
data is consistent with expected physical models of the volcano's activity.

\end{itemize}

This deployment was a significant learning experience, where many things went
wrong and many things were learned. Section~\ref{evaluation-sec-lessons}
describes some of the lessons in detail, linking this important feedback to
the architectural solutions in Chapters~\ref{chapter-lance} and
\ref{chapter-idea}.
