% Document template based on LNCS, adapted by Matt Welsh <mdw@cs.berkeley.edu>

% This version is adapted to use PDFTEX to render to PDF directly. 
% If you want to use dvi, you need to change any figures to use '.eps'
% rather than '.pdf', and probably get rid of the hyperref package.

%\documentclass[11pt, onecolumn, draftcls]{IEEEtran}
\documentclass[11pt]{IEEEtran}
\usepackage{program}
\usepackage{epsfig}
\usepackage{url}
\usepackage[english]{babel} % mdw: Required to get good hyphenation on RH6.0
                            % (fixed in RH6.1)
\usepackage{graphicx} 
\usepackage{color}    
\usepackage{cite}

\newcommand{\XXXnote}[1]{{\textcolor{red}{\bf MDW: #1}}}
\newcommand{\GWAnote}[1]{{\textcolor{red}{\bf GWA: #1}}}
%\newcommand{\XXXnote}[1]{}
%\newcommand{\GWAnote}[1]{}

\pagestyle{empty}
\begin{document}

%don't want date printed
\date{}

%%%%%%%%%%%% THIS IS WHERE WE PUT IN THE TITLE AND AUTHORS %%%%%%%%%%%%


% MDW: We need to say "Wireless Sensor Network" here
%\title{Developing a Wireless Sensor Network for High-Bandwidth, High-Fidelity
%Volcanic Monitoring}

\title{Deploying a Wireless Sensor Network on an Active Volcano}
\def\footnotemark{}

\author{\authorblockN{Geoffrey Werner-Allen\authorrefmark{1}, 
Konrad Lorincz\authorrefmark{1}, 
Mario Ruiz\authorrefmark{2}\authorrefmark{4}, 
Omar Marcillo\authorrefmark{3}\authorrefmark{4}, 
Jeff Johnson\authorrefmark{2},
Jonathan Lees\authorrefmark{3},
and Matt Welsh\authorrefmark{1}} \\
\authorblockA{\authorrefmark{1} Division of Engineering and Applied
Sciences, Harvard University} \\
\authorblockA{\authorrefmark{2} Dept. of Earth Sciences, University of
New Hampshire} \\
\authorblockA{\authorrefmark{3} Dept. of Geological Sciences, University of 
North Carolina} \\
\authorblockA{\authorrefmark{4} Instituto Geof\'{i}sico, Escuela 
Polit\'{e}cnica Nacional, Ecuador}}

%\author{\authorblockN{Geoffrey Werner-Allen, Konrad Lorincz, and Matt
%Welsh\\}
%\authorblockA{Harvard University, \{werner,konrad,mdw\}@eecs.harvard.edu\\}
%\authorblockN{Mario Ruiz and Jonathan Lees\\}
%\authorblockA{University of North Carolina, \{mruiz,leesj\}@email.unc.edu\\}
%\authorblockN{Omar Marcillo and Jeff Johnson\\}
%\authorblockA{University of New Hampshire,
%\{omar.marcillo,jeff.johnson\}@unh.edu}
%}

\maketitle
\thispagestyle{empty}

%%%%%%%%%%%%%  ABSTRACT GOES HERE %%%%%%%%%%%%%%

\begin{abstract} 
Wireless sensor networks have the potential to aid the study of active
volcanoes.  Augmenting heavy and power-hungry data collection
equipment with lighter, smaller wireless sensor network nodes leads to
faster, larger deployments.  Arrays consisting of dozens of wireless sensor
nodes are now possible.  The additional scale and resolution of these
networks permits scientific studies not feasible with traditional
instrumentation.  Designing sensor networks to support volcanic studies
requires addressing the high data rates and high data fidelity these studies
demand.  We have designed a sensor network application for volcanic data
collection that relies on triggered event detection and reliable data
retrieval to meet bandwidth and data quality demands.  Here we describe our
design and relate our experience deploying our network on Volc\'{a}n
Reventador, an active volcano in northern Ecuador.  
\end{abstract}

\input{intro.tex}
\input{description.tex}
\input{monitoring.tex}
\input{cs.tex}
\input{conclusions.tex}

%%%%%%%%%%%%%  THIS IS WHERE THE BIBLIOGRAPHY GOES %%%%%%%%%

\def\baselinestretch{0.92}
\begin{footnotesize}
\setlength{\itemsep}{2in}
\bibliographystyle{IEEEtran} \bibliography{IEEEComputing}
\end{footnotesize}
\end{document}
