\section{Motivation}
\label{lance-sec-motivation}

Wireless sensor networks are becoming more common for applications that focus
on reliable collection of raw signals at relatively high sampling rates, as
opposed to in-network aggregation of low-data-rate samples.  These
applications generally make use of extensive offline analysis to study the
collected data, and it is often infeasible or impossible to perform this
computation within the network itself.  Even in cases where it is possible to
shift computation to the network, a system designer may wish to extract raw
data occasionally for calibration or testing.  Examples of such applications
include structural health
monitoring~\cite{netshm-spots06,ggb-ipsn07,wimms-lynch06}, acoustic
sensing~\cite{vango,vigilnet,girod-ipsn07,enviromic}, distributed camera
networks~\cite{cyclops}, and geophysical monitoring~\cite{volcano-osdi06}.

These systems typically record data to flash at each sensor node and make use
of a reliable bulk-transfer protocol to collect data at a base station. Given
that the network is capable of sampling data at a higher rate than it can be
downloaded, it is not possible to collect the complete signal from all nodes.
The system is therefore forced to make a decision about what data to collect
and what to throw away. In most cases this decision is application-specific:
for example, a volcanologist may be chiefly interested in a specific type of
seismic tremor, and a biologist may be looking for acoustic signatures of a
specific species of marmot~\cite{girod-ipsn07}. The implication is that the
system must be able to determine the intrinsic {\em value} of a given signal
to determine whether resources should be devoted to storing and downloading
that signal.

Previous approaches have involved simple mechanisms tailored for specific
applications. For example, in the
NetSHM~\cite{netshm-spots06,netshm-emnets05} structural monitoring system,
data collection is triggered manually following an excitation of the
structure.  Sentri~\cite{ggb-ipsn07} has been used for vibration monitoring
at the Golden Gate Bridge; it is unclear from the paper how sampling and
communication are triggered, but reported experimental results suggest manual
operation. The Reventador volcano monitoring system~\cite{volcano-osdi06}
used a simple triggering algorithm to detect seismic events and initiate
reliable transfer to the base station.  The Intel Predictive Maintenance
system~\cite{intel-northseasensys} performs high-data-rate sampling staggered
over infrequent, periodic collection periods to extend system lifetime.  All
of these systems involve a tight binding of the {\em mechanisms} used to
manage storage and bandwidth with their respective application-specific {\em
policies}.

The typical approach to download management is a FIFO model where downloads
occur in a round-robin fashion across the network once a trigger occurs. In
general, new data may be sampled and stored to flash while a download is
taking place.  Therefore, the trigger frequency, download cycle duration and
the number of nodes in the network all effect the amount of data captured by
such an approach.  For example, the Flush~\cite{flush-sensys07} protocol
achieves only 8~Kbps for a reliable transfer over multiple hops; the
Fetch~\cite{volcano-osdi06} and Straw~\cite{ggb-ipsn07} protocols fare
somewhat worse. The RCRT protocol~\cite{rcrt-sensys07} is designed for a case
where all nodes are transmitting simultaneously to the sink, although this
approach severely limits the obtained per-node throughput. As a result, when
incoming data rates exceed download capacity, FIFO download management can
produce excessive delays between data acquisition and retrieval.

Lance assumes that sensor nodes contain adequate flash storage to buffer
signals prior to download.  While the popular TMote~Sky platform supports a
relatively small 1~MB flash, more recent sensor designs~\cite{shimmer} have
several GB~of flash, and we expect this trend to continue. Rather than
focusing on per-node storage, our primary concern is with limitations on
bandwidth and energy.

Our goal is to develop a general-purpose approach to bandwidth and energy
management that complements a reliable data-collection protocol. Such a
system should have several key properties.  First, it should be {\em
customizable}, allowing different applications to specify their own policies
for storage management and bandwidth prioritization. Second, the system
should target a range of optimization goals. Examples include maximizing
overall data priority, bounding energy consumption, maximizing temporal or
spatial coverage of the collected data, or achieving fairness across sensor
nodes. Third, the system should be decoupled from a specific routing
protocol, reliable collection protocol, or sensor node platform, making it
possible to leverage the system in different settings. 
