\documentclass[twocolumn,10pt]{usetex-v1}
\usepackage{graphicx,url,amsmath,amsfonts,amssymb,color,multirow}

\newcommand{\XXXnote}[1]{}
%\newcommand{\XXXnote}[1]{{\bf\textcolor{red}{XXX: #1}}}

\pagestyle{plain}
\pagenumbering{arabic}

\begin{document}

\title{IDEA: Integrated Distributed Energy Awareness\\for Wireless Sensor Networks}

\author{
  \authname{Geoffrey Werner Challen, Jason Waterman and Matt Welsh}
  \authaddr{School of Engineering and Applied Sciences}
  \authaddr{Harvard University}
  \authurl{\url{{challen,waterman,mdw}@eecs.harvard.edu}}
}

\maketitle
\thispagestyle{empty}

\subsection*{Abstract}
Energy in sensor networks is a distributed, non-transferable resource. Over
time, differences in energy availability are likely to arise. Protocols like
routing trees may concentrate energy usage at certain nodes. Differences in
energy harvesting arising from environmental variations, such as if one node
is in the sun and another is in the shade, can produce variations in charging
rates and battery levels. Because many sensor network applications require
nodes to collaborate --- to ensure complete sensor coverage or route data to
the network's edge --- a small set of nodes whose continued operation is
threatened by low batteries can have a disproportionate impact on the
fidelity provided by the network as a whole. In the most extreme case, the loss
of a single sink node may render the remainder of the network unreachable.

While previous research has addressed reducing the energy usage of individual
nodes, the challenge of collaborative energy management has been largely
ignored. We present Integrated Distributed Energy Awareness (IDEA), a sensor
network service enabling effective network-wide energy decision making. IDEA
\textit{integrates} into the sensor network application by providing an API
allowing components to evaluate their impact on other nodes.  IDEA
\textit{distributes} information about each node's load, charging, and
battery state to other nodes whose decisions affect it. Finally, IDEA enables
\textit{awareness} of the connection between the behavior of each node and
the application's energy goals, guiding the network toward states that
improve performance.

This paper describes the IDEA architecture and demonstrates its use through
three case studies. Using both simulation and testbed experiments, we
evaluate each IDEA application by comparing it to simpler approaches that do
not integrate distributed energy awareness. We show that using IDEA can
significantly improve performance compared with solutions operating with
purely local information.

\input{introduction.tex}
\input{motivation.tex}
\input{architecture.tex}
\input{casestudies.tex}
\input{evaluation.tex}
\input{related.tex}
\input{futurework.tex}

\begin{footnotesize}
\bibliographystyle{abbrv} \bibliography{mobisys10}
\end{footnotesize}

\end{document}
