% Document template based on LNCS, adapted by Matt Welsh <mdw@cs.berkeley.edu>

% This version is adapted to use PDFTEX to render to PDF directly. 
% If you want to use dvi, you need to change any figures to use '.eps'
% rather than '.pdf', and probably get rid of the hyperref package.

\documentclass[10pt]{article}
\usepackage{program}
\usepackage{acm-style10} % ACM proceedings formatting
\usepackage{times}  % Use Adobe Times font set
\usepackage{epsfig,twocolumn}
\usepackage{url}
\usepackage{textcomp}
\usepackage{usenix}
\usepackage[english]{babel} % mdw: Required to get good hyphenation on RH6.0
                            % (fixed in RH6.1)
\usepackage{graphicx} 
\usepackage{color}

% This will generate links within the PDF as well as a set of bookmarks
% \usepackage[colorlinks,pdfpagemode=None]{hyperref} 

%\def\dsp{\def\baselinestretch{0.92}\large\normalsize}
%\def\dsp{\def\baselinestretch{0.92}}
%\dsp
%\def\dsp{\def\baselinestretch{1.0}}
%\dsp

% DO NOT REMOVE THIS DEFINITION --- IF I HAVE LEFT AN XXXNOTE IN THE
% TEXT IT NEEDS TO BE ADDRESSED AND NOT HIDDEN!
\newcommand{\XXXnote}[1]{{\bf\color{red} XXX: #1}}
%\newcommand{\XXXnote}[1]{}

%\pagestyle{plain}
\pagestyle{empty}
%\pagenumbering{arabic}

%set dimensions of columns, gap between columns, and space between
%paragraphs
%\setlength{\textheight}{9.25in}
\setlength{\textheight}{9.0in}
\setlength{\columnsep}{0.25in}
\setlength{\textwidth}{6.75in}
\setlength{\footskip}{0.0in}
%\setlength{\topmargin}{-0.25in}
\setlength{\topmargin}{0.0in}
\setlength{\headheight}{0.0in}
\setlength{\headsep}{0.0in}
%\setlength{\oddsidemargin}{-0.15in}
\setlength{\oddsidemargin}{-0.125in}
\setlength{\evensidemargin}{0in}
\setlength{\parindent}{1pc}
\setlength{\itemindent}{0.0in}
%\setlength{\parskip}{\baselineskip}
\setlength{\baselineskip}{12pt}

%\setlength{\textheight}{9.25in}
%\setlength{\columnsep}{0.33in}
%\setlength{\textwidth}{7.4in}
%\setlength{\footskip}{0.0in}
%\setlength{\topmargin}{-0.25in}
%\setlength{\headheight}{0.0in}
%\setlength{\headsep}{0.0in}
%\setlength{\oddsidemargin}{-.45in}
%%\setlength{\oddsidemargin}{0in}
%\setlength{\parindent}{1pc}
%%\setlength{\parskip}{\baselineskip}

\begin{document}
% MDW: This needs to be HERE (after the \begin{document})
%\setlength{\baselineskip}{12pt}

%don't want date printed
\date{}

%%%%%%%%%%%% THIS IS WHERE WE PUT IN THE TITLE AND AUTHORS %%%%%%%%%%%%

\title{{\ttlfnt Fidelity and Yield in a \\ Volcano Monitoring Sensor
Network}}
\author{{\aufnt Geoff Werner-Allen$^\star$, Konrad Lorincz$^\star$, 
Jeff Johnson$^\dag$, Jonathan Lees$^\ddag$, and Matt Welsh$^\star$} \\
{\affaddr $^\star$ Division of Engineering and Applied Sciences, Harvard University} \\
{\affaddr $^\dag$ Dept. of Earth Sciences, University of New Hampshire} \\
{\affaddr $^\ddag$ Dept. of Geological Sciences, University of North
Carolina}}
\maketitle
%\copyrightspace

\thispagestyle{empty}

%%%%%%%%%%%%%  ABSTRACT GOES HERE %%%%%%%%%%%%%%

\subsection*{Abstract}
%\abstract{
\begin{small}
%\setlength{\baselineskip}{20pt}
We present a science-centric evaluation of a 19-day sensor
network deployment at Reventador, an active volcano in Ecuador.
Each of the 16~sensors continuously sampled seismic and acoustic 
data at 100~Hz. Nodes used an event-detection algorithm to trigger 
on interesting volcanic activity and initiate reliable data transfer 
to the base station. During the deployment, the network recorded
229~earthquakes, eruptions, and other seismoacoustic events. 
\end{small}

%\setlength{\baselineskip}{9pt}
\begin{small}
The science requirements of reliable data
collection, accurate event detection, and high timing precision 
drive sensor networks in new directions for geophysical monitoring.
The main contribution of this paper is an evaluation of the sensor
network as a scientific instrument, holding it to the standards
of existing instrumentation in terms of data {\em fidelity} (the
quality and accuracy of the recorded signals) and {\em yield} (the quantity
of the captured data).  
We describe an approach to {\em time rectification} of the acquired 
signals that can recover accurate timing despite failures of the 
underlying time synchronization protocol. In addition, we perform a 
detailed study of the sensor network's data using a direct comparison 
to a standalone data logger, as well as an investigation of seismic 
and acoustic wave arrival times across the network. 
%\setlength{\baselineskip}{9pt}
\end{small}
%}

%%%%%%%%%%%%%  BODY OF PAPER GOES HERE %%%%%%%%%%%%%%

\tensize
\setlength{\baselineskip}{12pt}

%\setlength{\baselineskip}{11.5pt}
\input{intro.tex} 
\input{background.tex}
\input{design.tex}
\input{robustness.tex}
\input{eventdetection.tex}
\input{performance.tex}
\input{timing.tex}
\input{fidelity.tex}
\input{related.tex}
%\input{conclusions.tex}
\input{lessons.tex}
\input{acknowledgements.tex}

%%%%%%%%%%%%%  THIS IS WHERE THE BIBLIOGRAPHY GOES %%%%%%%%%

\def\baselinestretch{0.92}
\begin{footnotesize}
\setlength{\itemsep}{1in}
\bibliographystyle{abbrv} \bibliography{osdi}
\end{footnotesize}

\end{document}
